---
format:
  revealjs:
    embed-resources: true
    slide-number: false
    progress: false
---

::: {style="text-align: center"}
##  How the Presence of Others<br>Can Bias Survey Responses {.center .smaller}

&nbsp;    

:::: columns

::: {.column width="33%"}

**Gustavo Diaz**  
McMaster University  
<diazg2@mcmaster.ca>  

:::

::: {.column width="33%"}
**Michelle Dion**  
McMaster University
<dionm@mcmaster.ca>
:::

::: {.column width="33%"}
**Guillem Riambau**  
Universitat de Barcelona 
Institut d'Economia de Barcelona  
<griambau@ub.edu>
:::

::::

&nbsp;

Slides and paper: [talks.gustavodiaz.org](https://talks.gustavodiaz.org/)

:::



```{r include=FALSE}
library(knitr)

opts_chunk$set(fig.pos = "center", echo = FALSE, 
               message = FALSE, warning = FALSE)
```

```{r setup}
# Packages
library(tidyverse)
library(DeclareDesign)
library(kableExtra)


# ggplot global options
theme_set(theme_gray(base_size = 20))
ggdodge = position_dodge(width = 0.5)
```

## Disclaimer {{< fa triangle-exclamation >}} {.center background-color="#FDBF57"}

Data harmonization still in progress!

## Sensitive survey questions

::: incremental
1. A social referent...

2. ... who *can learn* the response

3. ... has a *preferred* response 

4. Cost of *wrong* response
:::

::: aside
See [Tourangeau and Yan (2007)](https://psycnet.apa.org/doi/10.1037/0033-2909.133.5.859) and [Blair et al (2020)](https://doi.org/10.1017/S0003055420000374) for extensive discussions
:::

. . .

&nbsp;

{{< fa arrow-right >}} Presence of others meets all criteria

## What we know so far

::: incremental
- **Presence of enumerators:** Political questions are sensitive

- **Household surveys:** Presence of parents/spouses shapes answers

- **Afrobarometer:** Family and non-family affect responses and *don't know* rates [(Zimbalist 2022)](https://doi.org/10.1080/13645579.2021.1886397)
:::

## This project

1. More data and more details

2. Determine severity of problem

## This project

1. **More data and more details**

2. [Determine severity of problem]{style="color: gray;"}

## Asian Barometer Surveys

<!-- CONTINUE HERE DESCRIBE SURVEY AND WHY WE CHOOSE IT  -->

## Coverage

<!-- HERE SHOW PLOT WITH PRESENCE OF OTHERS ACROSS COUNTRIES  -->

## Sensitive questions

<!-- HERE SHOW LIST OF QUESTIONS WE FOCUS ON. REMEMBER TO MENTION THESE ARE NOT ALL THE QUESTIONS, AND THAT I WON'T SHOW EVERYTHING, JUST THE INTERESTING STUFF -->

## Results

<!-- PICK AND CHOOSE, DIFFERENT SLIDES FOR REFUSAL AND ANSWERS -->

## Conclusion

::: aside
**Learn more:** [talks.gustavodiaz.org](https://talks.gustavodiaz.org/)
:::
