---
format:
  revealjs:
    embed-resources: true
    slide-number: false
    progress: false
---

::: {style="text-align: center"}
##  How the Presence of Others<br>Can Bias Survey Responses {.center .smaller}

&nbsp;    

:::: columns

::: {.column width="33%"}

**Gustavo Diaz**  
McMaster University  
<diazg2@mcmaster.ca>  

:::

::: {.column width="33%"}
**Michelle Dion**  
McMaster University
<dionm@mcmaster.ca>
:::

::: {.column width="33%"}
**Guillem Riambau**  
Universitat de Barcelona 
Institut d'Economia de Barcelona  
<griambau@ub.edu>
:::

::::

&nbsp;

Slides and paper: [talks.gustavodiaz.org](https://talks.gustavodiaz.org/)

:::



```{r include=FALSE}
library(knitr)

opts_chunk$set(fig.pos = "center", echo = FALSE, 
               message = FALSE, warning = FALSE)
```

```{r setup}
# Packages
library(tidyverse)
library(haven)
library(DeclareDesign)
library(kableExtra)
library(vdemdata) # For electoral democracy scores
library(countrycode) # Convert country names to codes


# ggplot global options
theme_set(theme_gray(base_size = 20))
ggdodge = position_dodge(width = 0.5)
```

## Disclaimer {{< fa triangle-exclamation >}} {.center background-color="#FDBF57"}

Data harmonization still in progress!

## Sensitive survey questions

::: incremental
1. A social referent...

2. ... who *can learn* the response

3. ... has a *preferred* response 

4. Cost of *wrong* response
:::

::: aside
See [Tourangeau and Yan (2007)](https://psycnet.apa.org/doi/10.1037/0033-2909.133.5.859) and [Blair et al (2020)](https://doi.org/10.1017/S0003055420000374) for extensive discussions
:::

. . .

&nbsp;

{{< fa arrow-right >}} Presence of others meets all criteria

## What we know so far

::: incremental
- **Presence of enumerators:** Political questions are sensitive

- **Household surveys:** Presence of parents/spouses shapes answers

- **Afrobarometer:** Family and non-family affect responses and *don't know* rates [(Zimbalist 2022)](https://doi.org/10.1080/13645579.2021.1886397)
:::

## This project

1. More data and more details

2. Determine severity of problem

## This project

1. **More data and more details**

2. [Determine severity of problem]{style="color: gray;"}

## Asian Barometer Surveys

::: incremental

- Waves 1-5 (2000-2021)

- 1,000-1,800 respondents per country-wave

- 15 countries, 21 years, 97,439 interviews

- 95% Detailed record of who was present

- 45% of interviews had others present

- Partner/spouse > Children > Parents/in-laws > Nonfamily > Party/gov officials

:::

## Coverage

```{r}
ab = readRDS("privacy/ab.rds")

ab = ab %>% 
    mutate(year = ifelse(year < 2000 | year > 2021, NA, year))

ab = ab %>% 
    mutate(
        country = str_replace(country, "_", " "),
        country = str_to_sentence(country),
        country_id = countrycode(
        country,
        origin = "country.name",
        destination = "vdem"
    )) %>% 
    relocate(country_id)
```


```{r}
# Codebook: https://v-dem.net/documents/38/v-dem_codebook_v14.pdf
v = vdem %>% 
    select(country_name, country_id, year,
          v2x_libdem, v2x_regime, v2x_regime_amb) %>% 
    filter(year >= 2000 & year <= 2021)

abv = left_join(ab, v, by = c("country_id", "year"))

# binary democracy score
abv = abv %>% mutate(democracy = ifelse(v2x_regime <= 1, 0, 1))
```

```{r}
breakdown = abv %>% 
    summarize(record = mean(!is.na(bystander_any)),
              prop_present = mean(bystander_any, na.rm = TRUE),
              n = n(),
              .by = c(wave, country)
              ) %>% 
    mutate(prop_present = ifelse(record == 0, NA, prop_present))

ggplot(breakdown) +
    aes(x = as.factor(wave), y = country, 
        fill = as.factor(record)) +
    geom_tile(color = "white") +
    geom_text(aes(label = round(prop_present, 2)), size = 4) +
    scale_fill_grey(labels = c("Not recorded", "Recorded")) +
    scale_y_discrete(limits = rev) +
    theme(legend.position = "bottom") +
    labs(x = "Wave",
         y = "Country",
         fill = "Others present")
```



## Sensitive questions

::: incremental
1. Country is a democracy

2. Satisfaction with democracy

3. System of government is capable of solving country's problems

4. I would rather live under our system of government

5. Government will solve most important problem
:::

## Results: Refusal to answer
<!-- Anyone, Children, Parents/inlaws, neighbors/passersby -->

```{r, cache=TRUE}
refusal_outcomes = c("scale(this_country_a_democracy_NA_str)",
                     "scale(satisfaction_democracy_NA_str)",
                     "scale(system_capable_solving_NA_str)",
                     "scale(our_system_best_NA_str)",
                     "scale(gov_will_solve_problems_NA_str)")

# bystander_any
ref_any = refusal_outcomes %>% 
    map(~formula(paste0(., "~ bystander_any + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_any") %>% 
    relocate(outcome) %>% 
    mutate(group = "refusal")

# bystander_children
ref_children = refusal_outcomes %>% 
    map(~formula(paste0(., "~ bystander_children + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_children") %>% 
    relocate(outcome) %>% 
    mutate(group = "refusal")

# bystander_parent
ref_parent = refusal_outcomes %>% 
    map(~formula(paste0(., "~ bystander_parent + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_parent") %>% 
    relocate(outcome) %>% 
    mutate(group = "refusal")

# bystander_nonfamily
ref_nonfamily = refusal_outcomes %>% 
    map(~formula(paste0(., "~ bystander_nonfamily + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_nonfamily") %>% 
    relocate(outcome) %>% 
    mutate(group = "refusal")
```

```{r, cache=TRUE}
democracy_outcomes = c("scale(this_country_a_democracy_a)",
                     "scale(satisfaction_democracy_a)",
                     "scale(system_capable_solving_a)",
                     "scale(our_system_best_a)",
                     "scale(gov_will_solve_problems_a)")


# bystander_any
dem_any = democracy_outcomes %>% 
    map(~formula(paste0(., "~ bystander_any + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_any") %>% 
    relocate(outcome) %>% 
    mutate(group = "democracy")

# bystander_children
dem_children = democracy_outcomes %>% 
    map(~formula(paste0(., "~ bystander_children + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_children") %>% 
    relocate(outcome) %>% 
    mutate(group = "democracy")

# bystander_parent
dem_parent = democracy_outcomes %>% 
    map(~formula(paste0(., "~ bystander_parent + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_parent") %>% 
    relocate(outcome) %>% 
    mutate(group = "democracy")

# bystander_nonfamily
dem_nonfamily = democracy_outcomes %>% 
    map(~formula(paste0(., "~ bystander_nonfamily + woman + education_f + age_4_f + 
                        urban + religiosity_2_f"))) %>% 
    map(~lm_robust(., data = abv, se_type = "stata", weights = w,
                   fixed_effects = ~ country + wave)) %>% 
    map(tidy) %>%
    bind_rows() %>% 
    filter(term == "bystander_nonfamily") %>% 
    relocate(outcome) %>% 
    mutate(group = "democracy")
```


```{r}
main_df = rbind(ref_any, ref_children, ref_parent, ref_nonfamily,
      dem_any, dem_children, dem_parent, dem_nonfamily) %>% 
    mutate(outcome = str_remove(outcome, "_NA_str\\)"),
           outcome = str_remove(outcome, "scale\\("),
           outcome = str_remove(outcome, "_a\\)"))

# Reorder
main_df = main_df %>% 
    mutate(group = ifelse(group == "refusal", "Refusal", "Agreement"),
           group = fct_rev(group),
           term = fct_relevel(term,
                              "bystander_nonfamily",
                              "bystander_parent",
                              "bystander_children",
                              "bystander_any"),
           outcome = fct_relevel(outcome,
                                 "this_country_a_democracy",
                                 "satisfaction_democracy",
                                 "system_capable_solving",
                                 "our_system_best",
                                 "gov_will_solve_problems"))

ggplot(main_df %>% filter(group == "Refusal")) +
    aes(x = term, y = estimate, shape = outcome) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point(size = 2, position = position_dodge(width = 0.5)) +
    geom_linerange(aes(x = term, ymin = conf.low, ymax = conf.high),
                   position = position_dodge(width = 0.5)) +
    scale_x_discrete(labels = c("Nonfamily",
                                "Parents or in-laws",
                                "Children",
                                "Anyone")) +
    scale_shape_discrete(labels = c(
        "Country is\na democracy\n",
        "Satisfaction\nwith democracy\n",
        "System capable of\n solving country's problems\n",
        "Would rather live\nunder our system\n",
        "Gov. will solve\nmost important problem\n"
    )) +
    coord_flip() +
    # theme(legend.position = "bottom") +
    labs(x = "Presence of...",
         y = "Estimate",
         shape = "Outcome")
```


::: aside
Controls for sex, education, age, urban, religiosity. Country and wave fixed effects. HC2 standard errors.
:::

## Results: Agreement to questions

```{r}
ggplot(main_df %>% filter(group == "Agreement")) +
    aes(x = term, y = estimate, shape = outcome) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point(size = 2, position = position_dodge(width = 0.5)) +
    geom_linerange(aes(x = term, ymin = conf.low, ymax = conf.high),
                   position = position_dodge(width = 0.5)) +
    scale_x_discrete(labels = c("Nonfamily",
                                "Parents or in-laws",
                                "Children",
                                "Anyone")) +
    scale_shape_discrete(labels = c(
        "Country is\na democracy\n",
        "Satisfaction\nwith democracy\n",
        "System capable of\n solving country's problems\n",
        "Would rather live\nunder our system\n",
        "Gov. will solve\nmost important problem\n"
    )) +
    coord_flip() +
    # theme(legend.position = "bottom") +
    labs(x = "Presence of...",
         y = "Estimate",
         shape = "Outcome")
```


::: aside
Controls for sex, education, age, urban, religiosity. Country and wave fixed effects. HC2 standard errors.
:::


## Next steps

::: incremental
- Questions about trust, perceptions of corruption

- Determine whether presence of others biases statistical inferences

- **Takeaway:** Take into account in research design and analysis

:::

::: aside
**Learn more:** [talks.gustavodiaz.org](https://talks.gustavodiaz.org/)
:::
